Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-21 ms/main.tex
major reogranization of results, not finished yet

\autoref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}
%where $\hat{p}=v\tilde{p}+(1-v)p$.
\autoref{table:vars_params} summarizes the model variables and parameters.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Oblique and Horizontal Transmission}

With only oblique and horizontal transmission, i.e. $v = 0$, Eq.\ \ref{eq:horizontal} becomes $\hat{p}=p$ and Eq.\ \ref{eq:nextgen_adults_slimpify} becomes 
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) \;,
\end{equation}

which gives the following result.\\

\begin{result}[Oblique and horizontal transmission of cooperation] \label{result:obli_hori}
Without vertical transmission ($v=0$), if there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B \;,
\end{equation}
then the frequency of the cooperator phenotype among adults increases every generation, $p'>p$.
\end{result}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vertical and Horizontal Transmission}

With only vertical and horizontal transmission, i.e. $v=1$, Eq.\ \ref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and Eq.\ \ref{eq:nextgen_parents_simplified} for the frequency of the cooperative phenotype among parents in the next generation $\tilde{p}'$ can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

Fixation of either cooperation, $\tilde{p}=1$, or defection, 
$\tilde{p}=0$, are equilibria of Eq.\ \ref{eq:nextgen_parents_vertical_only}, solving $\tilde{p}'= \tilde{p}$.
We assume for the remainder of the analysis that $0<\tilde{p}<1$.

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ reduces to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0 \;,
\end{equation}
and there are no additional equilibria.
For cooperation to take over the population (i.e., for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$; that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} \;.
\end{equation}
Factoring out $\tilde{p}(1-\tilde{p})$ and setting $\bar{w} = 1 + \tilde{p}(b-c)$, we find that $\hat{p}'>\hat{p}$ if
%\begin{equation}
%  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
%  > 1 -\tilde{p} .
%\end{equation}
%Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A > 1 \;.
\end{equation}
\\

If $\alpha<1$, divide both sides of Eq.\ \ref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$. Then $\tilde{p}'>\tilde{p}$ if
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A \;.
\end{aligned}
\end{equation}
%Rearranging, we get
%\begin{equation} 
%\begin{aligned} 
%  1 - \tilde{p} < 
%  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
%  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
%  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
%  & + (1-\tilde{p})^2 (1-\alpha) T_A .
%\end{aligned}
%\end{equation}
%Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
%\begin{equation} 
%\begin{aligned} 
%  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
%   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
%   + (1-c) (1-\alpha) T_B
%   + (1+b) (1-\alpha) T_A 
%   - (1-\alpha) T_A].
%\end{aligned}
%\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B) \;.
\end{equation}

In addition to the the fixation states $\tilde{p}=0$ and $\tilde{p}=1$, there may be an actual polymorphic equilibrium of $\tilde{p}'=\tilde{p}$ in Eq.\ \ref{eq:nextgen_parents_vertical_only}, namely 
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B)} ,
\end{equation} 
which is legitimate if $0<\tilde{p}^*<1$.

Since all parameters are positive, we can apply
inequality \ref{eq:vert_hori_global_condition} and see that a requirement for $\tilde{p}'>\tilde{p}$ is that either 
\begin{align} 
\label{eq:vert_hori_global_condition1}
T_A > T_B \quad&\text{and}\quad \tilde{p}>\tilde{p}^*,  \quad \text{or} \\
\label{eq:vert_hori_global_condition2}
T_A <T_B \quad&\text{and}\quad \tilde{p}<\tilde{p}^* .
\end{align}

We define the \emph{cost boundaries},
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\gamma_1 = \frac{b \alpha T_A + (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}
\end{equation}
Then, applying Eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we  summarize the possible outcomes in the following result and corollaries.
\\

\begin{result}[Vertical and horizontal transmission of cooperation] \label{result:vert_hori}
With vertical and horizontal but without oblique transmission ($v=1$), the cultural evolution of cooperation exhibits one of the following scenarios, depending on the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}):

%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item \emph{Fixation of cooperation}:
	if \emph{(i)} $T_A \ge T_B$ and $c < \gamma_1$; or
	if \emph{(ii)} $T_A < T_B$ and $c < \gamma_2$.
\item \emph{Fixation of defection}:
	if \emph{(iii)} $T_A \ge T_B$ and $\gamma_2 < c$; or
	if \emph{(iv)} $T_A < T_B$ and $\gamma_1 < c$.
\item \emph{Stable coexistence}:
	if \emph{(v)} $T_A<T_B$ and $\gamma_2 < c < \gamma_1$.
\item \emph{Unstable coexistence}: 
	if \emph{(vi)} $T_A>T_B$ and $\gamma_1 < c <\gamma_2$.
\end{enumerate}
\end{result}

These conditions are illustrated in \autoref{fig:result2}.

Cooperation and defection can coexist stably at frequencies $\tilde{p}^*$ and $1-\tilde{p}^*$ (\autoref{eq:vert_hori_equilibrium}) if there is horizontal transmission bias for defection, $T_A<T_B$, and the cost of cooperation is intermediate, $\gamma_2 < c < \gamma_1$; see yellow area in \autoref{fig:result2}b.
When unstable coexistence occurs, phenotype $A$ will fix if its initial frequency is $p>\tilde{p}^*$, and phenotype $B$ will fix if its initial frequency is $1-p>1-\tilde{p}^*$; this occurs when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:coexistence_recursive}a shows the mapping $\tilde p \to \tilde p'$; see blue areas in \autoref{fig:result2}a and \autoref{fig:result2}b.
\\

\begin{corollary}[Symmetric horizontal transmission] 
\label{corollary:equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

The condition in inequality~\ref{eq:equal_transmission} demonstrates the importance of the social association $\alpha$. The following corollaries provide additional demonstrations.
\\

\begin{corollary}[No social association of transmission and cooperation]
\label{corollary:alpha0}
Without social association ($\alpha=0$), cooperation can take over the population if there is horizontal transmission bias for cooperation, $T_A>T_B$, and if either
\begin{equation}
\label{eq:vert_hori_alpha0}
c < \frac{T_A-T_B}{1-T_B}
\quad \text{or} \quad
\tilde{p}_0 > \tilde{p}^* = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B)} \;.
\end{equation}
\end{corollary}

\autoref{fig:result2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:vert_hori_equilibrium} and \ref{eq:cost_boundaries}. If there is horizontal transmission bias for cooperation ($T_A>T_B$) then cooperation can evolve from any initial frequency if the cost is low enough, $c<\gamma_1$ (green area below solid line). If the cost is not low enough, cooperation can also evolve if its initial frequency is high enough and the cost is not too high, $c<\gamma_2$ (blue area between solid and dashed lines).

We can interpret these condition as follows. First, when cooperators are rare, they will mostly interact with defectors. Therefore, for cooperation to increase in frequency and fix, the \emph{effective cost of cooperation} must be lower than the horizontal transmission of cooperation, $1-(1-c)(1-T_B) < T_A$, which is equivalent to the first of the two inequalities in \autoref{eq:vert_hori_alpha0}.
Second, if $T_A$ is not large enough, cooperation will fix if the initial frequency is higher than the threshold $\tilde{p}^*$; thus, this threshold must be below one, which entails $1-(1-c)(1-T_B) < T_A + b(T_A-T_B)$. That is, the effective cost of cooperation must be lower than the combined effect of horizontal transmission of cooperation, $T_A$, and the benefit of cooperation multiplied by the transmission bias, $b(T_A-T_B)$. 
\\

\begin{corollary}[Perfect social association of transmission and cooperation]
\label{corollary:alpha1}
With perfect social association ($\alpha=1$), the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This condition is obtained from inequality \ref{eq:vert_hori_alpha1_condition_proof}, and the right-hand side equals $\gamma_1$ when $\alpha=1$ (\autoref{eq:cost_boundaries}).
Perfect social association (horizontal transmission always occurs during the cooperative interaction) is also assumed in the model studied by \citet{lewin2017microbes}, and therefore this corollary is equivalent to the result in their eq.~1.

%In terms of the cost boundaries, inequality~\ref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$, and if $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, inequality~\ref{eq:vert_hori_alpha1} is sufficient for increase in the frequency of $A$.
Inequality~\ref{eq:vert_hori_alpha1} can also be written as
$1 - (1-c)(1-T_B) < (1+b) T_A$,
which provides an interesting interpretation for the success of cooperation. 
In the interaction between a cooperator and a defector, $(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, or that it fails to reproduce.
This is the \emph{effective cost of cooperation} from this interaction.
Similarly, $(1+b) T_A$ is the probability that during an interaction between a cooperator and a defector, the defector becomes cooperative and reproduces, which is the \emph{effective benefit of cooperation} from this interaction.
Thus, inequality~\ref{eq:vert_hori_alpha1} entails that cooperation can evolve if the effective cost of cooperation is less than the effective benefit during an interaction between a cooperator and a defector.
\\

Given the previous two corollaries for no association and perfect association ($\alpha=0$ and $\alpha=1$), it is interesting to examine the general effect of social association on the evolution of cooperation.
We denote the social association boundaries by
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting}
  a_1 = \frac{c(1-T_B)-(1+b)(T_A-T_B)}{b\cdot T_B}, \quad
  a_2 = \frac{c(1-T_B)-(T_A-T_B)}{b\cdot T_A}.
\end{aligned}\end{equation}
Applying Eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we have the following corollary.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]\label{corollary:intermediate_association}
Cooperation will increase from rarity if social association is high enough, specifically if $a_2 < \alpha$.
\end{corollary}

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{With Vertical and Oblique Transmission}

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
\autoref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
\autoref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
\autoref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
%TODO YR cleanup  
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

%We start with the simple case of symmetrical horizontal transmission, $T=T_A=T_B$ and apply \autoref{eq:horizontal}, \autoref{eq:nextgen_adults_slimpify}, and \autoref{eq:nextgen_parents_simplified} to obtain 
%\begin{equation} \label{eq:equal_horizontal_transmission}
%\begin{aligned}
%  f(\hat{p}) &= 
%  \bar{w}(\hat{p}' - \hat{p}) = \\
%  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].\end{aligned}
%\end{equation}
%A detailed explain of how we get eq.~\ref{eq:equal_horizontal_transmission} can be found in appendix B.
%The equilibria are solutions of $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
%It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
%Since the derivative of $f'(\hat{p}$ is 
%\begin{equation}
%f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
%\end{equation}
%then the condition for local stability of $\hat{p}=1$ is
%\begin{equation} \label{eq:derivative_of_phattag-phat}
%	f'(1) =	-\alpha bvT + cv(1-T) < 0,
%\end{equation}
%which gives the following result.
%
%\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}
%If horizontal transmission is symmetric, $T = T_A = T_B$, and if
%\begin{equation} \label{eq:oblique_and_vertic_result2}
%  %\frac{b}{c}>\frac{1-T}{\alpha T}
%	c < b \cdot \frac{\alpha T}{1-T},
%\end{equation}
%then fixation of the cooperator phenotype $A$ is locally stable.
%The same condition was given in Corollary 1.1, \autoref{eq:equal_transmission}.

%We now turn to the general case where $T_A \neq T_B$. 
Applying Eqs.\ \ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} gives the function $f(\hat{p})$ (see Appendix~\autoref{sec:appendixA}):

\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) , \\
\beta_2 &= -\beta_1 -\beta_3 ,  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1}.
\end{equation}

@@

Define new cost boundaries, $\hat\gamma_1$ and $\hat\gamma_2$,
\begin{equation} \label{eq:cost_boundaries_v}
\begin{aligned}
\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
\end{aligned}
\end{equation}
and  a vertical transmission threshold,
\begin{equation} \label{eq:v_threshold}
\hat v = \frac{T_B - T_A}{1-T_A} \,.
\end{equation}

@@

The following result summarizes the possible outcomes.
\\

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ (\autoref{eq:cost_boundaries_v}) and the vertical transmission threshold $\hat v$ (\autoref{eq:v_threshold}) :

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \hat\gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\hat\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\hat\gamma_1 < c$.
\item \emph{Stable Coexistence}: 
    if \emph{(v)} $T_A < T_B$ and $v<\hat{v}$ and $c < \hat\gamma_1$; or 
    if \emph{(vi)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_2 < c < \hat\gamma_1$.
\item \emph{Unstable coexistence}:
    if \emph{(vii)} $T_A > T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$.
\end{enumerate}

@@

Much of the literature on evolution of cooperation focuses on conditions for  initially rare cooperative phenotype to invade a population of defectors.
The next corollary deals with such a condition.
\\

\begin{corollary}[Condition for cooperation to increase from rarity]
\label{corollary:rarity}
  If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
  \begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
  \begin{aligned}
  c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
  \end{aligned}
  \end{equation} 
  \end{corollary}

>>>>

\autoref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}
%where $\hat{p}=v\tilde{p}+(1-v)p$.

Finally, we find an equation for the frequency of phenotype $A$ among juveniles in the next generation $\hat{p}'$ as a function of the frequency in the current generation.
Starting from \autoref{eq:horizontal}, we substitute \autoref{eq:nextgen_adults_slimpify} for $p'$ and  \autoref{eq:nextgen_parents_simplified} for $\tilde{p}'$.
We therefore have
\begin{equation} \label{eq:nextgen_juveniles}
  \begin{aligned}
  \hat{p}'  = & \frac{v}{\bar{w}}\left\{\hat{p}^2(1+b-c)\left[1-(1-\hat{p})(1-\alpha)T_B)\right]\right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1-c)\left[ \hat{p}(1-\alpha)T_B + 1 - T_B \right] \right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1+b)\left[\hat{p}(1-\alpha) + \alpha \right]T_A \right\} \\
  & + \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \\
  & + (1-v)\hat{p}^2(T_B-T_A) + (1-v)\hat{p}(1+T_A-T_B),
\end{aligned}
\end{equation}
where $\bar{w} = 1 + \hat{p}(b-c)$.

\autoref{table:vars_params} summarizes the model variables and parameters.

@@

In the following sections, we determine the equilibria of the model, namely, solutions of $\hat{p}' = \hat{p}$, and analyze their local stability.
We then analyze the evolution of a modifier of the social association to determine the evolution of social association.
Finally, we demonstrate the application of the conditions derived in the previous sections to predict the outcomes of stochastic simulations in a spatial model.

\subsection*{Equilibria and Stability}

To determine the equilibria in our model, we analyze the fixed points of \autoref{eq:nextgen_juveniles}.
We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics,  this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) , \\
\beta_2 &= -\beta_1 -\beta_3 ,  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$. 
Thus, $f(\hat{p})$ becomes a quadratic polynomial,
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixA}), where
$f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big]$, so that
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
%Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:general_equilibrium}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1}.
\end{equation}

@@

We define the \emph{cost boundaries}, $\hat\gamma_1$ and $\hat\gamma_2$,
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
\end{aligned}
\end{equation}
and the \emph{vertical transmission threshold}, $\hat v$,
\begin{equation} \label{eq:v_threshold}
\hat v = \frac{T_B - T_A}{1-T_A} \,.
\end{equation}

@@

The following result summarizes the possible outcomes of the model (\autoref{eq:nextgen_juveniles}).
\\

\begin{result}[Equilibria and stability] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ and the vertical transmission threshold $\hat v$ (Eqs.~\ref{eq:cost_boundaries}-\ref{eq:v_threshold}) :

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \hat\gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\hat\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\hat\gamma_1 < c$.
\item \emph{Stable coexistence}: 
    if \emph{(v)} $T_A < T_B$ and $v<\hat{v}$ and $c < \hat\gamma_1$; or 
    if \emph{(vi)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_2 < c < \hat\gamma_1$.
\item \emph{Unstable coexistence}:
    if \emph{(vii)} $T_A > T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$.
\end{enumerate}

@@

Much of the literature on evolution of cooperation focuses on conditions for  initially rare cooperative phenotype to invade a population of defectors.
The next remarks deal with such conditions. The first introduces a condition on the cost of cooperation, $c$, and the second on the social association, $\alpha$.
\\

\begin{remark}[Condition for cooperation to increase from rarity: cost of cooperation]
\label{corollary:rarity}
If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
\begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
\end{equation} 
\end{remark}

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection. In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.

Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation, $T_A$, increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), facilitating the evolution of cooperation.
However, increasing the horizontal transmission of defection, $T_B$, can increase or decrease the cost boundary, but it increases the cost boundary when it is already above one ($c<1<\hat\gamma_1$): $\partial\gamma_1/\partial T_B$ is positive when $T_A > \frac{1}{1+\alpha b v}$, which gives $\hat\gamma_1>1/v$. 
Therefore, increasing $T_B$ decreases the cost boundary and limits the evolution of cooperation, but only if $T_A < \frac{1}{1+\alpha b v}$.

Increasing the vertical transmission $v$ can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$, because $\text{sign}(\partial \hat\gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.
When $T_A<T_B$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the cost boundary increases, making it easier for cooperation to evolve from rarity. 
In contrast, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the cost boundary decreases making it harder for cooperation to evolve from rarity.
